---
title: "Harmonize Neonatal-Related Attributes in cMD Metadata"
author:
  - Britney Pheng, Sehyun Oh <br>
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
output: github_document
---

#### Load Packages
```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(dplyr)
})
```
```{r sampleMetadata, echo=FALSE}
data <- sampleMetadata

#class(sampleMetadata)
#dim(sampleMetadata)
#head(colnames(sampleMetadata), 10)
#sampleMetadata[1:4, 1:10]
```
#### Create a list of selected sampleMetadata columns to work with
```{r selected sampleMetadata columns}
selected_cols <- c("age_category",
                   "infant_age",
                   "born_method", 
                   "gestational_age",
                   "premature",
                   "birth_weight",
                   "lactating",
                   "c_section_type",
                   "feeding_practice",
                   "formula_first_day",
                   "breastfeeding_duration")
```
#### Create a new dataframe *nn.ds* with a new column for unique *study:sample_id* rows and the selected sampleMetadata columns
```{r dataframe, warning=FALSE}
nn.ds <- sampleMetadata %>%
    mutate(curation_id = paste(study_name, sample_id, sep = ":")) %>%
    dplyr::select(curation_id, selected_cols)
```
### Curated Neonatal Columns

#### Clean up *born_method* column into curated *neonatal_delivery_procedure* column
Use **c_section_type** data element to expand **born_method** column answers. 
C-section can be either an Emergency CS, Elective CS, or kept broadly as "C-section" if there is no expanded answer in the corresponding **c_section_type** column.
```{r curated delivery procedure}
nn.ds$neonatal_delivery_procedure <- NA

for (i in 1:nrow(nn.ds)) {
  if (is.na(nn.ds$born_method[i])) {
    nn.ds$neonatal_delivery_procedure[i] <- NA
  } else {
    if (nn.ds$born_method[i] == 'c_section' & !is.na(nn.ds$c_section_type[i])) {
    nn.ds$neonatal_delivery_procedure[i] <- nn.ds$c_section_type[i]
  } else {
      nn.ds$neonatal_delivery_procedure[i] <- nn.ds$born_method[i]
    }
  }
}
```
#### Clean up *premature* column into curated *neonatal_preterm_birth* column
Use **gestational_age** data element to clean up premature status. 

Pre-maturity (preterm birth) is defined as any birth less than 37 weeks and 0 days gestational age.
```{r curated preterm birth}
nn.ds$neonatal_preterm_birth <- NA

for (i in 1:nrow(nn.ds)) {
  if (is.na(nn.ds$gestational_age[i])) {
    nn.ds$neonatal_preterm_birth[i] <- nn.ds$premature[i]
  } else {
    if (nn.ds$gestational_age[i] < 37) {
      nn.ds$neonatal_preterm_birth[i] <- 'yes'
    } else if (nn.ds$gestational_age[i] >= 37) {
      nn.ds$neonatal_preterm_birth[i] <- 'no'
    } else {
      nn.ds$neonatal_preterm_birth[i] <- nn.ds$premature[x]
    }
  }
}
```
#### Clean up *feeding_practice* column into curated *neonatal_feeding_method* column
Replace "any_breastfeeding" **feeding_practice** data element answer with "exclusively_breastfeeding; mixed_feeding" answer.

Clean up "exclusively_breastfeeding" answer in the **feeding_practice** column to include any infant who was breastfed *and* did not receive formula between the time of birth and data collection. If **infant_age** data element is less than the day marked for **formula_first_day** element and there is a valid (non-NA) **breastfeeding_duration** answer, change feeding method answer to "exclusively_breastfeeding". This is with the notion that **infant_age** column provides the age of the infant at the time that samples were collected.
```{r curated feeding method}
nn.ds$neonatal_feeding_method <- NA

for (i in 1:nrow(nn.ds)) {
  if (is.na(nn.ds$feeding_practice[i])) {
    nn.ds$neonatal_feeding_method[i] <- nn.ds$feeding_practice[i]
  } else if (nn.ds$feeding_practice[i] == 'any_breastfeeding') {
    nn.ds$neonatal_feeding_method[i] <- 'exclusively_breastfeeding; mixed_feeding'
  } else if (is.na(nn.ds$formula_first_day[i]) == TRUE) {
    nn.ds$neonatal_feeding_method[i] <- nn.ds$feeding_practice[i]
  } else if (!(nn.ds$formula_first_day[i] <= nn.ds$infant_age[i]) 
             & !(is.na(nn.ds$breastfeeding_duration[i]))) {
    nn.ds$neonatal_feeding_method[i] <- 'exclusively_breastfeeding'
  } else {
    nn.ds$neonatal_feeding_method[i] <- nn.ds$feeding_practice[i]
  }
}
```
#### Rename unchanged neonatal data columns
```{r rename data columns, results='hide'}
nn.ds %>% 
  rename(
    neonatal_birth_weight = birth_weight,
    neonatal_gestational_age = gestational_age)
```
#### Generate a .csv file with the curated neonatal data columns
```{r create csv file}
write.csv(nn.ds, file = 'curated_neonatal_cols.csv')
```