---
title: "Harmonize Neonatal-Related Attributes in cMD Metadata"
author:
  - Britney Pheng, Sehyun Oh <br>
date: "`r format(Sys.time(), '%B %d, %Y')`"
format:
    html:
        fontsize: 14pxs
        toc: true
        top-depth: 3
output: github_document
---

# Setup
## Load packages
```{r install packages, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install('curatedMetagenomicData')
```

```{r}
suppressPackageStartupMessages({
  library(curatedMetagenomicData)
  library(tidyverse)
})
```

## Load data
```{r neonatal columns}
## A list of neonatal data column names
neonatal_cols <- c("infant_age",
                   "born_method", 
                   "gestational_age",
                   "premature",
                   "birth_weight",
                   "c_section_type",
                   "feeding_practice",
                   "formula_first_day",
                   "breastfeeding_duration")
```

```{r curated neonatal dataframe, warning=FALSE}
tb <- sampleMetadata %>%
    mutate(curation_id = paste(study_name, sample_id, sep = ":")) %>%
    dplyr::select(curation_id, neonatal_cols)
```

# neonatal_delivery_procedure
The two original columns `born_method` and `c_section_type` are harmonized to
the curated column, `neonatal_delivery_procedure`.

## EDA
```{r}
## 782 samples have information from both `born_method` and `c_section_type`
na_summary <- rowSums(!is.na(tb[c("born_method", "c_section_type")]))
table(na_summary)

## All the used terms
unique(tb$born_method)
unique(tb$c_section_type)
```
```{r}
both_ind <- which(na_summary == 2)
unique(tb$born_method[both_ind])

## No sample has only `c_section_type` info
sum(is.na(tb$born_method) & !is.na(tb$c_section_type)) 
```

## Consolidate redundant info
```{r curated delivery procedure}
tb$source <- NA
tb$value <- NA

for (i in seq_len(nrow(tb))) {
    if (is.na(tb$born_method[i])) {
        tb$source[i] <- NA
        tb$value[i] <- NA
    } else if (!is.na(tb$born_method[i]) & is.na(tb$c_section_type[i])) {
        tb$source[i] <- "born_method"
        tb$value[i] <- tb$born_method[i]
    } else {
        tb$source[i] <- "c_section_type"
        tb$value[i] <- tb$c_section_type[i]
    }
}
```

## Curate ontology terms
### Create curation map
The 'allowedvalues' and their 'ontology ids' are available in the data 
dictionary Google Sheet, `cMD_data_dictionary`, or in the GitHub repo.

You can look up the format of the curation maps in the manuscript's 
supplementary method section or other curation maps.

```{r}
delivery_map <- data.frame(
    original_value = c("Elective_CS", 
                       "Emergency_CS",
                       "c_section",
                       "vaginal"),
    curated_ontology_term = c("Elective Cesarean Delivery", 
                              "Emergency Cesarean Delivery",
                              "Cesarean Section",
                              "Vaginal Delivery"),
    curated_ontology_term_id = c("NCIT:C114141", 
                                 "NCIT:C92772", 
                                 "NCIT:C46088", 
                                 "NCIT:C81303"),
    curated_ontology_term_db = c("NCIT", "NCIT", "NCIT", "NCIT")
)
```
  
### Update the metadata with the curated terms 
```{r}
curated_delivery_dat <- tb %>%
    transmute(curation_id = curation_id,
              original_source = source,
              original_value = value,
              curated_ontology_term = plyr::mapvalues(
                  x = value, 
                  from = delivery_map$original_value,
                  to = delivery_map$curated_ontology_term,
                  warn_missing = FALSE
              )) %>%
    mutate(curated_ontology_term_id = plyr::mapvalues(
        x = curated_ontology_term,
        from = delivery_map$curated_ontology_term,
        to = delivery_map$curated_ontology_term_id,
        warn_missing = FALSE
    )) %>%
    mutate(curated_ontology_term_db = plyr::mapvalues(
        x = curated_ontology_term,
        from = delivery_map$curated_ontology_term,
        to = delivery_map$curated_ontology_term_db,
        warn_missing = FALSE
    ))
```

Check the harmonized/curated metadata
```{r}
non_na_ind <- which(!is.na(curated_delivery_dat$original_source))
head(curated_delivery_dat[c(1:3, non_na_ind),])
```

## Save the results
```{r}
readr::write_csv(delivery_map, "maps/cMD_delivery_procedure_map.csv")
readr::write_csv(curated_delivery_dat, "data/curated_delivery_procedure.csv")
```

# neonatal_preterm_birth
The two original columns `gestational_age` and `premature` are harmonized to the curated column `neonatal_preterm_birth`.

## EDA
```{r}
## 1891 samples have information for both `gestational_age` and `premature`
na_summary <- rowSums(!is.na(tb[c("gestational_age", "premature")]))
table(na_summary)
```

```{r}
## All the used values
unique(tb$premature)
sort(unique(tb$gestational_age), na.last=FALSE)
```

```{r}
## 329 samples have no `premature` info, but a provided `gestational_age` observation
sum(is.na(tb$premature) & !is.na(tb$gestational_age))

## 4 samples can have `neonatal_preterm_birth` == 'yes' 
sum(is.na(tb$premature) & tb$gestational_age < 37, na.rm=TRUE)

## 325 samples can have `neonatal_preterm_birth` == 'no'
sum(is.na(tb$premature) & tb$gestational_age >= 37, na.rm=TRUE)
```

## Consolidate redundant info
```{r}
tb$source <- NA
tb$value <- NA

for (i in seq_len(nrow(tb))) {
  if (is.na(tb$gestational_age[i])) {
    tb$source[i] <- NA
    tb$value[i] <- NA
  } else if (is.na(tb$premature[i]) & tb$gestational_age[i] < 37) {
    tb$source[i] <- "gestational_age"
    tb$value[i] <- "yes"
  } else if (is.na(tb$premature[i]) & tb$gestational_age[i] >= 37) {
    tb$source[i] <- "gestational_age"
    tb$value[i] <- "no"
  } else {
    tb$source[i] <- "premature"
    tb$value[i] <- tb$premature[i]
  }
}
```

## Curate ontology terms
### Create curation map

```{r}
premature_map <- data.frame(
  original_value = c("yes",
                     "no"),
  curated_ontology_term = c("Preterm",
                            "Term"),
  curated_ontology_term_id = c("NCIT:C92861",
                              "NCIT:C114093"),
  curated_ontology_term_db = c("NCIT", "NCIT")
)
```

### Update the metadata with the curated terms
```{r}
curated_premature_dat <- tb %>%
  transmute(curation_id = curation_id,
            original_source = source,
            original_value = value,
            curated_ontology_term = plyr::mapvalues(
              x = value,
              from = premature_map$original_value,
              to = premature_map$curated_ontology_term,
              warn_missing = FALSE
            )) %>%
    mutate(curated_ontology_term_id = plyr::mapvalues(
      x = curated_ontology_term,
      from = premature_map$curated_ontology_term,
      to = premature_map$curated_ontology_term_db,
      warn_missing = FALSE
    ))
```

Check the harmonized/curated metadata
```{r}
non_na_ind <- which(!is.na(curated_premature_dat$original_source))
head(curated_premature_dat[c(1:3, non_na_ind),])
```
### Save the results
```{r}
readr::write_csv(premature_map, "maps/cMD_preterm_birth_map.csv")
readr::write_csv(curated_premature_dat, "data/curated_preterm_birth.csv")
```

# neonatal_feeding_method
The three original columns `feeding_practice`, `formula_first_day`, and `breastfeeding_duration` are harmonized to the curated column `neonatal_feeding_method`.

## EDA
```{r}
## 136 samples have information for all three columns
na_summary <- rowSums(!is.na(tb[c("feeding_practice", "formula_first_day", "breastfeeding_duration")]))
table(na_summary)
```

```{r}
## All the used values
unique(tb$feeding_practice)

sort(unique(tb$formula_first_day), na.last=FALSE)

sort(unique(tb$breastfeeding_duration), na.last=FALSE)
```

```{r}
## No samples with a blank feeding practice and values for breastfeeding and formula feeding
sum(is.na(tb$feeding_practice) & !is.na(tb$formula_first_day) & !is.na(tb$breastfeeding_duration))

## 21 samples with "any_breastfeeding" value for `feeding_practice` variable
sum((tb$feeding_practice == "any_breastfeeding"), na.rm = TRUE)

## 85 samples where `feeding_practice` should be "exclusively_breastfeeding"
sum(!is.na(tb$breastfeeding_duration) & (tb$infant_age < tb$formula_first_day), na.rm = TRUE)

## 70 samples out of the 85 have "exclusively_breastfeeding" for `feeding_practice` column
sum((tb$feeding_practice == "exclusively_breastfeeding") & !is.na(tb$breastfeeding_duration) & (tb$infant_age < tb$formula_first_day), na.rm = TRUE)

## 15 samples should have "exclusively_breastfeeding" as their feeding_practice value given that the first day of formula feeding was after the date the sample was collected
sum((tb$feeding_practice == 'mixed_feeding') & !is.na(tb$breastfeeding_duration) & (tb$infant_age < tb$formula_first_day))
```

## Consolidate redundant info
```{r}
tb$source <- NA
tb$value <- NA

for (i in seq_len(nrow(tb))) {
  if (is.na(tb$feeding_practice[i])) {
    tb$source[i] <- NA
    tb$value[i] <- NA
  } else if (tb$feeding_practice[i] == "any_breastfeeding") {
    tb$source[i] <- "feeding_practice"
    tb$value[i] <- "exclusively_breastfeeding; mixed_feeding"
  } else if ((tb$formula_first_day[i] > tb$infant_age[i]) &
             !is.na(tb$breastfeeding_duration[i]) &
             !(tb$feeding_practice[i] == "exclusively_breastfeeding")) {
    tb$source[i] <- "formula_first_day"
    tb$value[i] <- "exclusively_breastfeeding"
  } else {
    tb$source[i] <- "feeding_practice"
    tb$value[i] <- tb$feeding_practice[i]
  }
}
```

## Curate ontology terms
### Create curation map
"mixed_feeding"               "exclusively_breastfeeding"  
[4] "no_breastfeeding"            "any_breastfeeding"           "exclusively_formula_feeding"

```{r}
feeding_map <- data.frame(
  original_value = c("exclusively_breastfeeding; mixed_feeding",
                     "exclusively_breastfeeding",
                     "exclusively_formula_feeding",
                     "mixed_feeding",
                     "no_breastfeeding"
                     ),
  curated_ontology_term = c("Exclusively Breastfeeding; Mixed Feeding",
                            "Exclusively Breastfeeding",
                            "Exclusively Formula Feeding",
                            "Mixed Feeding",
                            "No Breastfeeding"),
  curated_ontology_term_id = c("ONS:1000029; NA",
                              "ONS:1000029",
                              "MAXO:0000775",
                              NA,
                              NA),
  curated_ontology_term_db = c("ONS; NA", "ONS", "MAXO", NA, NA)
)
```

### Update the metadata with the curated terms
```{r}
curated_feeding_dat <- tb %>%
  transmute(curation_id = curation_id,
            original_source = source,
            original_value = value,
            curated_ontology_term = plyr::mapvalues(
              x = value,
              from = feeding_map$original_value,
              to = feeding_map$curated_ontology_term,
              warn_missing = FALSE
            )) %>%
    mutate(curated_ontology_term_id = plyr::mapvalues(
      x = curated_ontology_term,
      from = feeding_map$curated_ontology_term,
      to = feeding_map$curated_ontology_term_db,
      warn_missing = FALSE
    ))
```

Check the harmonized/curated metadata
```{r}
non_na_ind <- which(!is.na(curated_feeding_dat$original_source))
head(curated_feeding_dat[c(1:3, non_na_ind),])
```
### Save the results
```{r}
readr::write_csv(feeding_map, "maps/cMD_feeding_map.csv")
readr::write_csv(curated_feeding_dat, "data/curated_feeding_method.csv")
```